\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-11 paper/ms.tex
edits

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2J-J_{real}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

We initialize \exampleivnwalkers~walkers by sampling from an isotropic
Gaussian centered on the MAP parameters (the full set of parameters and their
priors are listed in Table~\ref{tab:example-4-params}), run
\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
\emcee.
We estimate the mean autocorrelation time for the chains of
$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
samples from the marginalized posterior density.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
This result is consistent within the published error bars and the posterior
constraints are tighter than the published results.
\Figure{astero_psd} shows a comparison of
the power spectrum estimated with a Lomb-Scargle periodogram 
with the power spectrum inferred with \celerite.  Despite the much smaller 
one-month sequence of data used in the \celerite\ analysis, the power spectrum 
inferred with \celerite\ matches well the periodogram estimated from
the full four years of Kepler data.
All asteroseismic analyses are known to have substantial systematic and
method-dependent uncertainties \citep{Verner:2011} so further experiments
would be needed to fully assess the reliability of this specific model.

@@

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector $\bvec{q}$ of random normal deviates, and then multiplying
\begin{equation}
\bvec{y} = L D^{1/2} \bvec{q}.
\end{equation}
This may be carried out in $\mathcal{O}(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets as even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
and multiplication are used.

@@

Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
\begin{eqnarray}
\bar f^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\left(a_j \cos{(d_j \vert t^*_m-t_n\vert)} +b_j \sin{(d_j \vert t^*_m-t_n\vert)}\right) z_j,
\end{eqnarray}
for $m =1$ to $M$.
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
$\{t_{n0+1},...,t_N\} \ge t^*_m$, gives
\begin{eqnarray}
\bar f^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t^*_m-t_n)} \left[a_j \cos{(d_j(t^*_m-t_n))}+b_j\sin{(d_j(t^*_m-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t^*_m)} \left[a_j \cos{(d_j(t_n-t^*_m))}+b_j\sin{(d_j(t_n-t^*_m))}\right] z_n.
\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{n,i} &=& \left(Q^-_{n-1,i}+z_n\tilde V_{n,i}\right)e^{-c_j(t_{n+1}-t_{n})},\\
X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i}^*,\\
Q^+_{n,i} &=& \left(Q^+_{n+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_{n}-t_{n-1})},\\
X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i}^*,
\end{eqnarray}
where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluted at $t_m^*$.
Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
point the predicted mean can be computed from
\begin{equation}
\bar f^*_i = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
\end{equation}

>>>>

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2\,J-J_{real}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

We initialize \exampleivnwalkers~walkers by sampling from an isotropic
Gaussian centered on the MAP parameters (the full set of parameters and their
priors are listed in Table~\ref{tab:example-4-params}), run
\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
\emcee.
We estimate the mean autocorrelation time for the chains of
$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
samples from the marginalized posterior density.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
This result is consistent within the published error bars and the posterior
constraints are tighter than the published results.
The top two panels of \Figure{astero_psd} show the Lomb-Scargle periodogram
\citep{VanderPlas:2017} estimator for the power spectrum based on the full
4~years of \kepler\ data and the month-long subset used in our analysis.
The bottom panel of \Figure{astero_psd} shows the posterior inference of the
PSD using the \celerite\ and the model described here.
Despite only using one month of data, the \celerite\ inference captures the
salient features of the power spectrum and it is qualitatively consistent with
the standard estimator applied to the full baseline.
All asteroseismic analyses are known to have substantial systematic and
method-dependent uncertainties \citep{Verner:2011} so further experiments
would be needed to fully assess the reliability of this specific model.

@@

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector $\bvec{q}$ of random normal deviates, and then multiplying
\begin{equation}
\bvec{y} = L D^{1/2} \bvec{q}.
\end{equation}
This may be carried out in $\mathcal{O}(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2\,J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2\,J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets as even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
and multiplication are used.

@@

Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
\begin{eqnarray}
\bar f^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\left(a_j \cos{(d_j \vert t^*_m-t_n\vert)} +b_j \sin{(d_j \vert t^*_m-t_n\vert)}\right) z_j,
\end{eqnarray}
for $m =1$ to $M$.
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
$\{t_{n0+1},...,t_N\} \ge t^*_m$, gives
\begin{eqnarray}
\bar f^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t^*_m-t_n)} \left[a_j \cos{(d_j(t^*_m-t_n))}+b_j\sin{(d_j(t^*_m-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t^*_m)} \left[a_j \cos{(d_j(t_n-t^*_m))}+b_j\sin{(d_j(t_n-t^*_m))}\right] z_n.
\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{n,i} &=& \left(Q^-_{n-1,i}+z_n\tilde V_{n,i}\right)e^{-c_j(t_{n+1}-t_{n})},\\
X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i}^*,\\
Q^+_{n,i} &=& \left(Q^+_{n+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_{n}-t_{n-1})},\\
X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i}^*,
\end{eqnarray}
where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$
to $2\,J$, and in $X^\pm_{m,n,i}$ the expressions
for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluted at $t_m^*$.
Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
point the predicted mean can be computed from
\begin{equation}
\bar f^*_i = \sum_{i=1}^{2\,J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
\end{equation}
